Replacing standard galaxy profiles with \\ mixtures of Gaussians

### 2013-03-10 documents/mixtures/mixture_models.tex
working through email comments

We are not the first in this space: Deconvolution and modeling of
galaxy images with MoGs has been done very successfully before (for
example, \citealt{bendinelli, emsellem, bendinelli2, cappellari}; they
use the name ``multi-Gaussian expansion'' or MGE).  However, the idea
behind those projects was to use the MoGs to provide a very free form
for the isophotes or morphologies of resolved galaxies or other
complex scenes.  Here our goals are very limited: We want to improve
the performance of standard galaxy image model fitting by expressing
the standard galaxy models---the exponential and de~Vaucouleurs
profiles---as rigid MoGs.

@@

Not only does a sampling representation of the profiles require many
more compute operations per render than a MoG representation, it also
requires more code complexity.  For the sampling to be both accurate
and efficient, it has to be strongly adaptive (with much higher
density in the center).  This adds complexity and choices into the
code, leading to engineering and software maintenance and testing
overheads that cost not just compute time but human time.  One
response to this is to build large numbers of pre-computed profile PSF
convolutions and interpolate among them.  This is the strategy taken
in the \project{SDSS} pipelines (\citealt{lupton}), but that requires
even \emph{more} human overhead, because not only does the sampling
and convolution have to be done correctly, so does the numerical
interpolation.  Finally, a sampling approximation---or any finite
numerical approximation---is effectively something like a ``mixture of
delta functions'' or ``mixture of integration kernels'' approximation.
Any such approximation (at fixed accuracy) will have far more degrees
of freedom (complexity) than the MoG approximations given below.

@@

We need good performance in two-dimensional image synthesis (for
fitting) because, with \thetractor\ (Lang \etal, forthcoming), we are
building a comprehensive model of all the imaging data we have; this
will have models of many millions of galaxies in imaging that contains
on the order of $10^{13}$ pixels.  We are basing our models on the
\project{Sloan Digital Sky Survey} (\sdss) Catalog galaxy models,
which only include exponential, de~Vaucouleurs, and composite (mixture
of the two) radial profiles.  In detail, in fact, the \sdss\ Catalog
models are small modifications of these; details below.  So our goal
here is to provide replacements for these models in order to improve
the performance of galaxy image modeling and analysis software.  Some
of the models we present have been used previously in tools for
precise photometry (\citealt{bundy}), and are being used in the Large
Synoptic Survey Telescope (LSST) prototype galaxy photometry pipeline
(\citealt{shaw}), in which these models are convolved with a shapelet
representation of the PSF; because shapelets are simply polynomial
perturbations on a Gaussian, they can also be convolved analytically
with mixture-of-Gaussian galaxy models (\citealt{bosch}).  These
models could also be used to speed other image-fitting systems, like
the very successful \project{GALFIT} (\citealt{galfit}).  An
alternative to making MoG profiles is to find exact analytic
expressions for certain kinds of convolutions.  There \emph{are} some
analytic results for PSF-convolved S\'ersic models but they involve
some non-trivial series and special-function expressions
(\citealt{trujillo}); we will not come back to these again but they
might be very useful in many situations.

@@

The results of the optimizations are shown in \tablename
s~\ref{tab:exp} and \ref{tab:lux} and \figurename s~\ref{fig:exp}
through \ref{fig:M}.  All the results shown in these figures and
tables and more are available in machine-readable form from author DWH
upon request.  In the \tablename s and \figurename s we show
root-variances rather than variances because these have units of
half-light radii; they are simple standard deviations for the Gaussian
components.

@@

One amusing aspect of MoG profiles has to do with projection from
three to two dimensions.  The projection of a three-dimensional
Gaussian is a two-dimensional Gaussian; the two-dimensional, rigid,
circular approximations we have made for the ser profiles can be
deprojected to rigid, spherical approximations to the
three-dimensional profiles trivially.  The two-dimensional models we
have started with are not accurate models of galaxies in detail---no
galaxy follows exactly any ser profile---so deprojection of our
approximations are not that interesting in themselves.  However,
the general program of fitting two-dimensional sources with MoGs may
have strong implications in the future for three-dimensional modeling
and deprojection.

@@

\begin{thebibliography}{70}
\bibitem[Bendinelli(1991)]{bendinelli}
Bendinelli,~O., 1991, \apj, 366, 599
\bibitem[Bendinelli \& Parmeggiani(1995)]{bendinelli2}
Bendinelli,~O. \& Parmeggiani,~G., 1995, \aj, 109, 572
\bibitem[Bosch(2010)]{bosch}
Bosch,~J., 2010, \aj, 140, 870
\bibitem[Bovy \etal(2011a)]{xd}
Bovy,~J., Hogg,~D.~W., \& Roweis,~S., 2011, Ann. Appl. Stat., 5, 1657
\bibitem[Bovy \etal(2011b)]{xdqso}
Bovy,~J., \etal, 2011, \apj, 729, 141
\bibitem[Bovy \etal(2012)]{xdqsoz}
Bovy,~J. \etal, 2012, \aj, 749, 41
\bibitem[Bundy \etal(2012)]{bundy}
Bundy,~K. \etal, \aj, in press
\bibitem[Cappellari(2002)]{cappellari}
Cappellari,~M., 2002, \mnras, 333, 400
\bibitem[Ciotti \& Bertin(1999)]{ciotti}
Ciotti,~L. \& Bertin,~G., 1999, \aa, 352, 447
\bibitem[de~Vaucouleurs(1948)]{dev}
de~Vaucouleurs,~G., 1948, Annales d'Astrophysique, 11, 247
\bibitem[Emsellem \etal(1994)]{emsellem}
Emsellem,~E., Monnet,~G., Bacon,~R., \& Nieto,~J.-L., 1994, \aap, 285, 739
\bibitem[Lupton \etal(2001)]{lupton}
Lupton,~R., Gunn,~J.~E., Ivezic,~Z., Knapp,~G.~R., Kent,~S.~M., \& Yasuda,~N., 2001, ASPC, 238, 269
\bibitem[Peng \etal(2002)]{galfit}
Peng,~C.~Y., Ho, L.~C., Impey,~C.~D., \& Rix,~H.-W., 2002, \aj, 124, 266
\bibitem[S{\'e}rsic(1963)]{ser}
S{\'e}rsic,~J.~L., 1963, Boletin de la Asociacion Argentina de Astronomia La Plata Argentina, 6, 41
\bibitem[Shaw(2012)]{shaw}
Shaw,~R.~A., ed., 2012, LSST Data Challenge Handbook (Version 2.0; Tucson, AZ: LSST Corp.)
\bibitem[Simard et al.(2011)]{bulgedisk}
Simard,~L., Mendel,~J.~T., Patton,~D.~R., Ellison,~S.~L., \& McConnachie,~A.~W., 2011, \apjs, 196, 11
\bibitem[Trujillo \etal(2001)]{trujillo}
Trujillo,~I., Aguerri,~J.~A.~L., Cepa,~J., \& Guti{\'e}rrez,~C.~M., 2001, \mnras, 321, 269 
\end{thebibliography}

>>>>

We are not the first in this space: Deconvolution and modeling of
galaxy images with MoGs has been done very successfully before (for
example, \citealt{bendinelli, emsellem, bendinelli2, kochanek, cappellari}; they
use the name ``multi-Gaussian expansion'' or MGE).  However, the idea
behind those projects was to use the MoGs to provide a very free form
for the isophotes or morphologies of resolved galaxies or other
complex scenes.  Here our goals are very limited: We want to improve
the performance of standard galaxy image model fitting by expressing
the standard galaxy models---the exponential and de~Vaucouleurs
profiles---as rigid MoGs.

@@

Not only does a sampling representation of the profiles require many
more compute operations per render than a MoG representation, it also
requires more code complexity.  For the sampling to be both accurate
and efficient, it has to be strongly adaptive (with much higher
density in the center).  This adds complexity and choices into the
code, leading to engineering and software maintenance and testing
overheads that cost not just compute time but human time.  One
response to this is to build large numbers of pre-computed profile PSF
convolutions and interpolate among them.  This is the strategy taken
in the \project{SDSS} pipelines (\citealt{lupton}), but that requires
even \emph{more} human overhead, because not only does the sampling
and convolution have to be done correctly, so does the numerical
interpolation.  Another less common strategy is to find other kinds of
function expansions (\citealt{bendi}).  Finally, a sampling
approximation---or any finite numerical approximation---is effectively
something like a ``mixture of delta functions'' or ``mixture of
integration kernels'' approximation.  Any such approximation (at fixed
accuracy) will have far more degrees of freedom (complexity) than the
MoG approximations given below.

@@

We need good performance in two-dimensional image synthesis (for
fitting) because, with \thetractor\ (Lang \etal, forthcoming), we are
building a comprehensive model of all the imaging data we have; this
will have models of many millions of galaxies in imaging that contains
on the order of $10^{13}$ pixels.  We are basing our models on the
\project{Sloan Digital Sky Survey} (\sdss) Catalog galaxy models,
which only include exponential, de~Vaucouleurs, and composite (mixture
of the two) radial profiles.  In detail, in fact, the \sdss\ Catalog
models are small modifications of these; details below.  Our goal
here is to provide replacements for these models in order to improve
the performance of galaxy image modeling and analysis software.  Some
of the models we present have been used previously in tools for
precise photometry (\citealt{bundy}), and are being used in the Large
Synoptic Survey Telescope (LSST) prototype galaxy photometry pipeline
(\citealt{shaw}), in which these models are convolved with a shapelet
representation of the PSF; because shapelets are simply polynomial
perturbations on a Gaussian, they can also be convolved analytically
with mixture-of-Gaussian galaxy models (\citealt{bosch}).  These
models could also be used to speed other image-fitting systems, like
the very successful \project{GALFIT} (\citealt{galfit}).  An
alternative to making MoG profiles is to find exact analytic
expressions for certain kinds of convolutions.  There \emph{are} some
analytic results for PSF-convolved S\'ersic models but they involve
some non-trivial series and special-function expressions
(\citealt{trujillo}); we will not come back to these again but they
might be very useful in many situations.

@@

The results of the optimizations are shown in \tablename
s~\ref{tab:exp} and \ref{tab:lux} and \figurename s~\ref{fig:exp}
through \ref{fig:M}.  All the results shown in these \figurename s and
\tablename s and more are available in machine-readable form from author DWH
upon request.  In the \tablename s and \figurename s we show
root-variances rather than variances because these have units of
half-light radii; they are simple standard deviations for the Gaussian
components.

@@

One amusing aspect of MoG profiles has to do with projection from
three to two dimensions.  The projection of a three-dimensional
Gaussian is a two-dimensional Gaussian; the two-dimensional, rigid,
circular approximations we have made for the ser profiles can be
deprojected to rigid, spherical approximations to the
three-dimensional profiles trivially.  The two-dimensional models we
have started with are not accurate models of galaxies in detail---no
galaxy follows exactly any ser profile---so deprojection of our
approximations are not that interesting in themselves.  However, the
general program of fitting two-dimensional sources with MoGs may have
strong implications in the future for three-dimensional modeling and
deprojection; some work has been done in this direction previously
(\citealt{vdb}); it is promising.

@@

\begin{thebibliography}{70}
\bibitem[Bendinelli(1991)]{bendinelli}
Bendinelli,~O., 1991, \apj, 366, 599
\bibitem[Bendinelli et al.(1993)]{bendi}
Bendinelli,~O., Ciotti,~L., \& Parmeggiani,~G.\ 1993, \aap, 279, 668 
\bibitem[Bendinelli \& Parmeggiani(1995)]{bendinelli2}
Bendinelli,~O. \& Parmeggiani,~G., 1995, \aj, 109, 572
\bibitem[Bosch(2010)]{bosch}
Bosch,~J., 2010, \aj, 140, 870
\bibitem[Bovy \etal(2011a)]{xd}
Bovy,~J., Hogg,~D.~W., \& Roweis,~S., 2011, Ann. Appl. Stat., 5, 1657
\bibitem[Bovy \etal(2011b)]{xdqso}
Bovy,~J., \etal, 2011, \apj, 729, 141
\bibitem[Bovy \etal(2012)]{xdqsoz}
Bovy,~J. \etal, 2012, \aj, 749, 41
\bibitem[Bundy \etal(2012)]{bundy}
Bundy,~K. \etal, \aj, in press
\bibitem[Cappellari(2002)]{cappellari}
Cappellari,~M., 2002, \mnras, 333, 400
\bibitem[Ciotti \& Bertin(1999)]{ciotti}
Ciotti,~L. \& Bertin,~G., 1999, \aa, 352, 447
\bibitem[de~Vaucouleurs(1948)]{dev}
de~Vaucouleurs,~G., 1948, Annales d'Astrophysique, 11, 247
\bibitem[Emsellem \etal(1994)]{emsellem}
Emsellem,~E., Monnet,~G., Bacon,~R., \& Nieto,~J.-L., 1994, \aap, 285, 739
\bibitem[Kochanek et al.(2000)]{kochanek}
Kochanek,~C.~S., Falco,~E.~E., Impey,~C.~D., et al., 2000, \apj, 535, 692
\bibitem[Lupton \etal(2001)]{lupton}
Lupton,~R., Gunn,~J.~E., Ivezic,~Z., Knapp,~G.~R., Kent,~S.~M., \& Yasuda,~N., 2001, ASPC, 238, 269
\bibitem[Peng \etal(2002)]{galfit}
Peng,~C.~Y., Ho, L.~C., Impey,~C.~D., \& Rix,~H.-W., 2002, \aj, 124, 266
\bibitem[S{\'e}rsic(1963)]{ser}
S{\'e}rsic,~J.~L., 1963, Boletin de la Asociacion Argentina de Astronomia La Plata Argentina, 6, 41
\bibitem[Shaw(2012)]{shaw}
Shaw,~R.~A., ed., 2012, LSST Data Challenge Handbook (Version 2.0; Tucson, AZ: LSST Corp.)
\bibitem[Simard et al.(2011)]{bulgedisk}
Simard,~L., Mendel,~J.~T., Patton,~D.~R., Ellison,~S.~L., \& McConnachie,~A.~W., 2011, \apjs, 196, 11
\bibitem[Trujillo \etal(2001)]{trujillo}
Trujillo,~I., Aguerri,~J.~A.~L., Cepa,~J., \& Guti{\'e}rrez,~C.~M., 2001, \mnras, 321, 269 
\bibitem[van~den~Bosch et al.(2008)]{vdb}
van~den~Bosch,~R.~C.~E., van de Ven,~G., Verolme,~E.~K., Cappellari,~M., \& de~Zeeuw,~P.~T., 2008, \mnras, 385, 647
\end{thebibliography}
